\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2017-05-05 doc/publications/plos16/nm.tex
added some of Oscars chnages

The most popular random forest software packages to date are R-based, although it has been reported that current implementations 
do not scale well with increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
than $2^{31}-1$), and the \randomforest implementation in R lending itself to parallelisation by growing multiple
trees simultaneously and combining them for the final result~\cite{Liaw.and.Weiner.2002}.  More suitable implementations for ``wide''
genomic data have been developed in other programming languages. \textsc{CloudForest} \cite{Bressler2015}, written in Go,
achieves fast run times by effective use of the CPU cache, optimising for different classes of features and efficiently
multithreading.  Similarly, \ranger~\cite{Wright.and.Ziegle.2016}, written in C++ with an R front-end, is
specifically optimised for dealing with large data sets.

However, the use of traditional compute infrastructure limits the parallelisation strategies that can be employed.  
The programs are limited to utilising only CPUs that are on the same computer node (multithreading) or
farm out independent tasks to CPUs distributed across nodes that do not require communication between the
processes (a separate tree is grown on each node).
%else they can run on CPUs distributed across nodes by virtue of the fact that there is no communication between the
%processes (a separate tree is grown on each node).  
Hadoop/Spark overcomes these limitations by enabling programs to
scale beyond compute-node boundaries and hence enable more sophisticated parallelisation strategies.  In the case of
random forests, the computations for each node of a tree can hence be handed off to separate processors.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This section talks about the data sets
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Data Sets}
\subsubsection{1000 Genomes Project}
We obtained the 1000 Genomes Project data as VCF files from their FTP site.  Each VCF file contains the variants for
every individual across one chromosome.  We use the phase 3 dataset which contains 2,504 individuals with approximately
81 million variant positions (i.e. a matrix of $2,504 \times 81$ million).  For compatibility with the majority of the
algorithms we demonstrate, we convert the original variant representations to double representations.  For example, we
store the absence of a variant (0\textbar0) as 0, a heterozygous
variant (1\textbar0 or 0\textbar1) as 1, and a
homozygous variant (1\textbar1) as 2, so we have an ``additive'' model.  However, this is not the only possible encoding \cite{Goldstein.et.al.2011}. Note that from
the encodings listed in Table~\ref{table:encodings}, the additive encoding will present the most difficulty for a random
forest model.

>>>>

The most popular random forest software packages to date are implemented in R \cite[]{R-language}, although it has been
reported that current implementations do not scale well with increasing feature size~\cite{Wright.and.Ziegle.2016}.
This is despite R now supporting long vectors (length greater than $2^{31}-1$), and the \randomforest implementation in
R lending itself to parallelisation by growing multiple trees simultaneously and combining them for the final
result~\cite{Liaw.and.Weiner.2002}.  More suitable implementations for ``wide'' genomic data have been developed in
other programming languages. \textsc{CloudForest} \cite{Bressler2015}, written in Go, achieves fast run times by
effective use of the CPU cache, optimising for different classes of features and efficiently multithreading.  Similarly,
\ranger~\cite{Wright.and.Ziegle.2016}, written in C++ with an R front-end, is specifically optimised for dealing with
large data sets.

However, the use of traditional compute infrastructure limits the parallelisation strategies that can be employed.  
The programs are limited to utilising only CPUs that are on the same computer node (multithreading) or
farm out independent tasks to CPUs distributed across nodes that do not require communication between the
processes (a separate tree is grown on each node).
%else they can run on CPUs distributed across nodes by virtue of the fact that there is no communication between the
%processes (a separate tree is grown on each node).  
Hadoop/Spark overcomes these limitations by enabling programs to
scale beyond compute-node boundaries and hence enable more sophisticated parallelisation strategies. In the case of
random forests, the computations for each node of a tree can hence be handed off to separate processors (cores).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This section talks about the data sets
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Data Sets}
\subsubsection{1000 Genomes Project}
We obtained the 1000 Genomes Project data as VCF files from their FTP site \cite[]{1KG2012}.  Each VCF file contains the variants for
every individual across one chromosome.  We use the phase 3 dataset which contains 2,504 individuals with approximately
81 million variant positions (i.e. a matrix of $2,504 \times 81$ million).  For compatibility with the majority of the
algorithms we demonstrate, we convert the original variant representations to double representations.  For example, we
store the absence of a variant (0\textbar0) as 0, a heterozygous
variant (1\textbar0 or 0\textbar1) as 1, and a
homozygous variant (1\textbar1) as 2, so we have an ``additive'' model.  However, this is not the only possible encoding \cite{Goldstein.et.al.2011}. Note that from
the encodings listed in Table~\ref{table:encodings}, the additive encoding will present the most difficulty for a random
forest model.
